### 2025-05-29 mss.tex
Post

\documentclass[twoside]{article}

\usepackage{ustj}

\usepackage{tikz}
\newcommand{\bit}[1]{%
  \tikz[baseline=(char.base)]{%
    \node[shape=rectangle, draw, rounded corners=1pt, inner sep=1pt, minimum width=6pt, minimum height=6pt] (char) {#1};
  }%
}

\setcounter{tocdepth}{2}

\addbibresource{mss.bib}

\newcommand{\authorname}{N. E. Davis}
\newcommand{\authorpatp}{\patp{lagrev-nocfep}}
\newcommand{\affiliation}{Zorp Corp}

%  Make first page footer:
\fancypagestyle{firststyle}{%
\fancyhf{}% Clear header/footer
\fancyhead{}
\fancyfoot[L]{{\footnotesize
              %% We toggle between these:
              % Manuscript submitted for review.\\
              {\it Urbit Systems Technical Journal} II:1 (2025):  1–46. \\
              ~ \\
              Address author correspondence to \authorpatp.
              }}
}
%  Arrange subsequent pages:
\fancyhf{}
\fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
\fancyhead[RO]{Representing Nouns as Byte Arrays}
\fancyfoot[LE,RO]{\thepage}

%%MANUSCRIPT
\title{Representing Nouns as Byte Arrays}
\author{\authorname~\authorpatp \\ \affiliation}
\date{}

\begin{document}

\maketitle
\thispagestyle{firststyle}

\begin{abstract}
  Noun serialization is commonly used for Nock communication, both between instances like Urbit ships and with the runtime and the Unix host operating system.  This article describes the two principal conventions for representing nouns in slightly compressed form as byte arrays, as well as introduces a new na\"ive run-length encoding for educational purposes.
\end{abstract}

% We will adjust page numbering in final editing.
\pagenumbering{arabic}
\setcounter{page}{1}

\tableofcontents

Nock deals in nouns and does not know about bits or memory.  However, a Nock interpreter (runtime) must deal with these practicalities.  In other words, we must have a way of writing down an abstract binary tree (consisting of cells/pairs and atoms) as an actual, physical array of bits in memory.

There are three basic ways to encode a noun as an atom:

\begin{enumerate}
  \item  \texttt{++jel}.  A na\"{i}ve encoding similar to \texttt{++jam} but without references.  Introduced in this article.
  \item  \texttt{++jam}.  A run-length encoding including references to repeated values.  The standard serialization for Urbit/Arvo.
  \item  \texttt{++bulk}/\texttt{++silo}.  A directed graph encoding with references.  The standard serialization for Nockchain.
\end{enumerate}

\noindent{}
Each of these

\subsection{\texttt{++jel}, the Na\"{i}ve Encoding}

The simplest way to encode a noun as a binary tree in an atom (byte array) without compression is to utilize bits to mark atom (\texttt{0}) vs. cell (\texttt{1}), the length of the atom in unary of \texttt{1}s terminated by \texttt{0}, and the actual value.  (Since the leading zeros are stripped, they may need to be added back in.)  This is in least-significant \emph{byte} (\textsc{lsb}) order, so you should read the written atom starting from the \emph{right}.

Thus the atom \texttt{0b0} would simply be encoded as:

\begin{lstlisting}[style=listingcode]
:: 0x0 = 0 for atom; 1 for length (special case);
::       0 for end of length; 0 for value
> `@ub`(jel 0b0)
0b10
\end{lstlisting}

\noindent{}
interpreted as (rightmost, LSB) \texttt{0} for atom, run length of \texttt{1}, and value \texttt{0} (stripped).  I.e., \texttt{0b010} or

{
\centering
$\leftarrow$ \\
\bit{0} \bit{1} \bit{0}
}

Other values include:

\begin{lstlisting}[style=listingcode]
:: 0x1 = 0 for atom; 1 for length;
::       0 for end of length; 1 for value
> `@ub`(jel 0b1)
0b1010

:: [0x0 0x1] = 1 for cell; zero, then one
> `@ub`(jel [@0b0@ 0b1])
0b1.010@0.010@1

:: [0x1 0x0] = 1 for cell; one, then zero
> `@ub`(jel [@0b0@ 0b1])
0b10@1.010@1

:: 0x2 = 0 for atom; 2 for length;
::       0 for end of length; 2 for value
> `@ub`(jel 0b10)
0b10.0111

:: 0xff
> `@ub`(jel 0xff)
0b11.1111.1101.1111.1110

:: [[0x0 0x1] [0x2 0x3]]
> `@ub`(jel [@0b0@ 0b1])
0b10@1.010@1
\end{lstlisting}

Our code implementation for \texttt{++jel} is as follows:

\begin{lstlisting}[style=listingcode]
!:  |%
++  jel
  =jel !:  |=  a=*
  ^-  @
  =+  l=0
  =+  b=0
  =<  -
  |-
  ?^  a
    =+  lv=$(a -.a)
    =+  rv=$(a +.a)
    =+  [c l]=(mash rv lv)
    [(con (lsh [0 1] c) 0b1) +(l)]
  ?:  =(0 a)  [0b10 4]
  :: need another mash in here for unary length
  =+  [c l]=(mash [a l] [b +((met 0 b))])
  [(con (lsh [0 1] c) 0b0) +(l)]
:: length of atom in unary
++  len
  |=  a=@
  ^-  @
  (fil 0 (met 0 a) 0b1)
:: mash two atoms together
++  mash
  |=  [a=[p=@ l=@] b=[p=@ l=@]]
  ^-  [c=@ l=@]
  :-  (con (lsh [0 l.b] p.a) p.b)
  (add l.a l.b)
--
\end{lstlisting}

Note that \texttt{l} is not the length of an atom in unary, but the length of the encoded noun in binary.

\subsection{\texttt{++jam}, the Practical Serialization}

The \texttt{++jel} algorithm has two flaws:  it is subject to collisions XXX
and it is verbose.  \texttt{++jam} will improve on this by altering the \textsc{rle} slightly and

special-cases 0

The new \textsc{rle} calculation is to post the number plus one in binary rather than unary (e.g., for \texttt{} the length would be \texttt{0b010}).

(Since there is a value less significant than the length, the leading zero is not lost but serves as a divider.)

\begin{lstlisting}[style=listingcode]
++  jam
  ~/  %jam
  |=  a=*
  ^-  @
  =+  b=0
  =+  m=`(map * @)`~
  =<  q
  |-  ^-  [p=@ q=@ r=(map * @)]
  =+  c=(~(get by m) a)
  ?~  c
    =>  .(m (~(put by m) a b))
    ?:  ?=(@ a)
      =+  d=(mat a)
      [(add 1 p.d) (lsh 0 q.d) m]
    =>  .(b (add 2 b))
    =+  d=$(a -.a)
    =+  e=$(a +.a, b (add b p.d), m r.d)
    :+  (add 2 (add p.d p.e))
      (mix 1 (lsh [0 2] (cat 0 q.d q.e)))
    r.e
  ?:  ?&(?=(@ a) (lte (met 0 a) (met 0 u.c)))
    =+  d=(mat a)
    [(add 1 p.d) (lsh 0 q.d) m]
  =+  d=(mat u.c)
  [(add 2 p.d) (mix 3 (lsh [0 2] q.d)) m]
\end{lstlisting}

### 2025-05-29 mss.tex
Post newt

@@
 
-\newcommand{\authorname}{N. E. Davis}
-\newcommand{\authorpatp}{\patp{lagrev-nocfep}}
-\newcommand{\affiliation}{Zorp Corp}
+\newcommand{\authorname}{N. E. Davis \& Sam Parker}
+\newcommand{\authorpatp}{\patp{lagrev-nocfep} \& \patp{zod}}
+\newcommand{\affiliation}{Zorp Corp \& —}
 
@@
 
-\subsection{\texttt{++jel}, the Na\"{i}ve Encoding}
+\section{\texttt{++jel}, the Na\"{i}ve Encoding}
 
@@
 
-\subsection{\texttt{++jam}, the Practical Serialization}
+\section{\texttt{++jam}, the Practical Serialization}
 
@@
 \end{lstlisting}
+
+\subsection{\texttt{newt} Encoding}
+
+“Newt” encoding is a runtime-oriented extension of \texttt{++jam}-based noun serialization which adds a short identifying header in case of future changes to the serialization format.
+A version number (currently as a single bit) precedes a serialization length followed by the \texttt{++jam} serialization of the noun.
+The version number is currently always \texttt{0b0}, but this may change in the future.
+
+\begin{lstlisting}[style=listingcode]
+  V.LLLL.JJJJ.JJJJ.JJJJ.JJJJ.JJJJ.JJJJ
+\end{lstlisting}
+
+\noindent{}
+where \texttt{V} is the version number, \texttt{L} is the total length of the noun in bytes, and \texttt{J} is the \texttt{++jam} serialization of the noun.
+
+Runtime communications vanes like \texttt{\%khan} and \texttt{\%lick} utilize this encoding locally.  It is exclusively used as a Unix affordance at the current time.
+
+\section{\texttt{+\$bulk}, the Directed Graph Encoding}
+
+\end{document}

### 2025-10-19 mss.tex
In media res

--- 2025-05-29 mss.tex
@@
 
-\newcommand{\authorname}{N. E. Davis}
-\newcommand{\authorpatp}{\patp{lagrev-nocfep}}
-\newcommand{\affiliation}{Zorp Corp}
+\newcommand{\authorname}{N. E. Davis, Brian Klatt, \& Sam Parker}
+\newcommand{\authorpatp}{\patp{lagrev-nocfep}, \patp{zod}, \& \patp{zod}}
+\newcommand{\affiliation}{Zorp Corp, Zorp Corp, \& —}
 
@@
               %% We toggle between these:
-              % Manuscript submitted for review.\\
-              {\it Urbit Systems Technical Journal} II:1 (2025):  1–46. \\
-              ~ \\
-              Address author correspondence to \authorpatp.
+              Manuscript submitted for review.\\
+              % {\it Urbit Systems Technical Journal} II:2 (2025):  1–46. \\
+              % ~ \\
+              % Address author correspondence to \authorpatp.
               }}
@@
 \fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
-\fancyhead[RO]{Representing Nouns as Byte Arrays}
+\fancyhead[RO]{Serializing Nouns}
 \fancyfoot[LE,RO]{\thepage}
@@
 %%MANUSCRIPT
-\title{Representing Nouns as Byte Arrays}
-\author{\authorname~\authorpatp \\ \affiliation}
+\title{Serializing Nouns}
+\author{N.\ E.\ Davis \patp{lagrev-nocfep}, Brian Klatt \patp{zod},\thanks{Brian Klatt contributed the description of \texttt{++jam}.} \\ \& Sam Parker \patp{zod}\thanks{Sam Parker contributed the description of \texttt{++bulk}.} \\ \affiliation}
+% TODO Brian Klatt credit
 \date{}
@@
 Noun serialization is commonly used for Nock communication, both between instances like Urbit ships and with the runtime and the Unix host operating system.
-This article describes the two principal conventions for representing nouns in slightly compressed form as byte arrays, as well as introduces a new na\"ive run-length encoding for educational purposes.
+This article describes and compares the two principal conventions for representing nouns in slightly compressed form as byte arrays, as well as introduces variant encodings for educational purposes.
 \end{abstract}
@@
 
-Nock deals in nouns and does not know about bits or memory.
-However, a Nock interpreter (runtime) must deal with these practicalities.
-In other words, we must have a way of writing down an abstract binary tree (consisting of cells/pairs and atoms) as an actual, physical array of bits in memory.
+\section{Introduction}
 
-There are three basic ways to encode a noun as an atom:
+The Nock combinator calculus deals in nouns and does not know about bit encodings or memory layouts.
+A Nock interpreter (runtime) must, however, deal with the practicalities.
+In other words, there must be a way of writing down an abstract binary tree (consisting of cells/pairs and atoms) as an actual, physical array of bits in memory.
+Every possible Nock noun can be represented as a finite sequence of bytes (an atom), and there are multiple ways to do so.\footnote{The converse is not true:  not every atom represents a valid deserialization or conversion from a graph encoding.}
+
+A noun serialization strategy is rather like a Gödel numbering in that it systematically encodes a mathematical object (a noun) as a number (an atom).
+Unlike Gödel numbering, which classically serially encodes the symbols of mathematical statements, noun serialization encodes a binary tree structure.\footnote{This also echoes the way that S-expressions are encoded in Lisp.}  Because a noun may be any atom---and atoms cannot have leading zeroes---both structure and value need to be unambiguously encoded and cannot be simply delimited (as by a \texttt{0} bit or similar).
+There are two basic strategies to encode a noun as an atom:
 
 \begin{enumerate}
-  \item  \texttt{++jel}.  A na\"{i}ve encoding similar to \texttt{++jam} but without references.  Introduced in this article.
-  \item  \texttt{++jam}.  A run-length encoding including references to repeated values.  The standard serialization for Urbit/Arvo.
-  \item  \texttt{++bulk}/\texttt{++silo}.  A directed graph encoding with references.  The standard serialization for Nockchain.
+  \item  Run-length serialization, with or without references.
+  \item  Directed graph serialization, depending on a reentrant graph encoding.
 \end{enumerate}
 
-\noindent{}
-Each of these
+\noindent
+Both of these embody a tradeoff between simplicity, size, and speed.  In this article, we describe both of these strategies and some new variants which offer possible advantages.
 
-\subsection{\texttt{++jel}, the Na\"{i}ve Encoding}
+\section{Na\"{i}ve Serialization}
 
@@
 cell (\texttt{1}), the length of the atom in unary of \texttt{1}s terminated by \texttt{0}, and the actual value.
-(Since the leading zeros are stripped, they may need to be added back in.)  This is in least-significant \emph{byte} (\textsc{lsb}) order, so you should read the written atom starting from the \emph{right}.
+(Since the leading zeros are stripped, they may need to be added back in.)  This is in least-significant \emph{byte} (\textsc{lsb}) order, so you should read the written atom starting from the \emph{right} in binary.
 
@@
 
-\noindent{}
+\noindent
 interpreted as (rightmost, LSB) \texttt{0} for atom, run length of \texttt{1}, and value \texttt{0} (stripped).  I.e., \texttt{0b010} or
@@
 
-\subsection{\texttt{++jam}, the Practical Serialization}
+\section{Practical Serialization}
 
-The \texttt{++jel} algorithm has two flaws:  it is subject to collisions XXX
-and it is verbose.  \texttt{++jam} will improve on this by altering the \textsc{rle} slightly and
+Whatever the pedagogical advantages of \texttt{++jel}, the algorithm has practical flaws:  it is subject to collisions XXX
+and it is verbose.\footnote{Note the claim of \patp{dozreg-toplud}, p.~TODO of this issue, that an operational Arvo instance may have up to $1.66 \times 10^{21}$ nouns, reduced by structural sharing.}  \texttt{++jam} improves the basic strategy by altering the \textsc{rle} algorithm slightly and supporting internal references for noun subtrees that have already been encoded.
+
+\texttt{++jam} converts a noun into a buffer and deduplicates repeated subtrees.
+It walks subtrees and encodes each in a way that allows for efficient storage and retrieval, while also permitting references to previously encoded values.
+
+- get Bryan notes
 
@@
 \end{lstlisting}
+
+A Python example of \texttt{++jam} is included in Appendix~A.
+
+\subsection{\texttt{newt} Encoding}
+
+\sloppy
+“Newt” encoding is a runtime-oriented extension of \texttt{++jam}\-based noun serialization which adds a short identifying header in case of future changes to the serialization format.
+A \mbox{version} number (currently a single bit) precedes a \textsc{rle} serialization length followed by the \texttt{++jam} serialization of the noun.
+The version number is currently \texttt{0b0}.
+
+\begin{lstlisting}[style=listingcode]
+  V.LLLL.JJJJ.JJJJ.JJJJ.JJJJ.JJJJ.JJJJ
+\end{lstlisting}
+
+\noindent
+where \texttt{V} is the version number, \texttt{L} is the total length of the noun in bytes, and \texttt{J} is the \texttt{++jam} serialization of the noun.
+
+Runtime communications vanes like \texttt{\%khan} and \texttt{\%lick} utilize this encoding locally.  It is exclusively used as a host OS runtime affordance at the current time.
+
+\section{Directed Graph Encoding}
+
+A directed graph encoding has been independently proposed twice, once by Tlon in the original Hoon codebase as a \lstinline[style=inlinecode]{+$silo} encoding and once by Sam Parker as the \lstinline[style=inlinecode]{++bulk} encoding.
+
+\section{Aligned Serialization}
+
+One of the advantages of \texttt{++jam} is its compactness.
+However, this comes at the cost of speed, since bit-level operations are required to \texttt{++cue} the noun back from its serialized form.
+If a slightly larger size is acceptable, a byte-aligned serialization could facilitate certain kinds of external inspection without requiring deserialization.
+(For instance, a byte-aligned head tag could be read for a rapid decision without needing to \texttt{++cue} the entire noun.)
+
+We propose a strategy to modify \texttt{++jam} to align to bytes by padding the length of entries to the nearest byte boundary and marking the distance with a clever binary scheme rather than simply unary.
+This approach, called \texttt{++honey}, aims to balance compactness and speed for certain use cases while retaining a large degree of conceptual backwards compatibility.
+(The change in byte alignment of course breaks strict compatibility.)
+
+There are two fundamental issues for byte alignment:  atoms and lengths.
+Atoms can be padded with leading zeros to the nearest byte boundary without changing their value.
+Lengths, however, require a new encoding scheme to compensate for the adjustment in expected bit widths.
+%We propose a variable-length encoding for lengths that uses the high bit of each byte to indicate whether more bytes follow.
+The remaining seven bits of each byte contribute to the length value.
+This allows for efficient representation of lengths while maintaining byte alignment.
+
+\section{Benchmarks}
+
+\section{Conclusion}
+
+\section*{Appendix A:  Python \texttt{++jam}/\texttt{++cue}}
+
+The following is a simple Python implementation of \texttt{++jam} serialization drawn from Urbit’s auxiliary \texttt{pynoun} library.
+
+\begin{lstlisting}[style=listingcode_python]
+from bitstring import BitArray
+noun = int | Cell
+# The Cell class represents an ordered pair of two nouns.
+
+def jam_to_stream(n: noun, out: BitArray):
+    """jam but put the bits into a stream
+
+    >>> s = BitArray()
+    >>> jam_to_stream(Cell(0,0), s)
+    >>> s
+    BitArray('0b100101')
+    """
+
+    cur = 0
+    refs = {}
+
+    def bit(b: bool):
+        nonlocal cur
+        out.append([b])
+        cur += 1
+
+    def zero():
+        bit(False)
+
+    def one():
+        bit(True)
+
+    def bits(num: int, count: int):
+        nonlocal cur
+        for i in range(0, count):
+            out.append([(num & (1 << i)) != 0])
+        cur += count
+
+    def save(a: noun):
+        refs[a] = cur
+
+    def mat(i: int):
+        if 0 == i:
+            one()
+        else:
+            a = i.bit_length()
+            b = a.bit_length()
+            above = b + 1
+            below = b - 1
+            bits(1 << b, above)
+            bits(a & ((1 << below) - 1), below)
+            bits(i, a)
+
+    def back(ref: int):
+        one()
+        one()
+        mat(ref)
+
+    def r(a: noun):
+        dupe = refs.get(a)
+        if deep(a):
+            if dupe:
+                back(dupe)
+            else:
+                save(a)
+                one()
+                zero()
+                r(a.head)
+                r(a.tail)
+        elif dupe:
+            isize = a.bit_length()
+            dsize = dupe.bit_length()
+            if isize < dsize:
+                zero()
+                mat(a)
+            else:
+                back(dupe)
+        else:
+            save(a)
+            zero()
+            mat(a)
+    r(n)
+
+def jam(n: noun):
+    """urbit serialization: * -> @
+
+    >>> jam(0)
+    2
+    >>> jam(Cell(0,0))
+    41
+    >>> jam(Cell(Cell(1234567890987654321, \\
+    ...               1234567890987654321), \\
+    ...          Cell(1234567890987654321, \\
+    ...               1234567890987654321)))
+    22840095095806892874257389573
+    """
+
+    out = BitArray()
+    jam_to_stream(n, out)
+    return read_int(len(out), out)
+
+def cue_from_stream(s: BitArray):
+    """cue but read the bits from a stream
+
+    >>> s = BitArray('0b01')
+    >>> cue_from_stream(s)
+    0
+    """
+
+    refs = {}
+    cur = 0
+    position = 0
+
+    def bits(n: int):
+        nonlocal cur, position
+        cur += n
+        result = 0
+        for i in range(n):
+            result |= (1 if s[position] else 0) << i
+            position += 1
+        return result
+
+    def one():
+        nonlocal cur, position
+        cur += 1
+        bit = s[position]
+        position += 1
+        return bit
+
+    def rub():
+        z = 0
+        while not one():
+            z += 1
+        if 0 == z:
+            return 0
+        below = z - 1
+        lbits = bits(below)
+        bex = 1 << below
+        return bits(bex ^ lbits)
+
+    def r(start: int):
+        ret = None
+        if one():
+            if one():
+                ret = refs[rub()]
+            else:
+                hed = r(cur)
+                tal = r(cur)
+                ret = Cell(hed, tal)
+        else:
+            ret = rub()
+        refs[start] = ret
+        return ret
+    return r(cur)
+
+def cue(i: int):
+    """urbit deserialization: @ -> *
+
+    >>> str(cue(22840095095806892874257389573))
+    '[[1234567890987654321 1234567890987654321]
+      1234567890987654321 1234567890987654321]'
+    """
+
+    bits = BitArray()
+    while i > 0:
+        bits.append([i & 1 == 1])
+        i >>= 1
+    return cue_from_stream(bits)
+\end{lstlisting}
+
+\end{document}

### 2025-11-01 mss.tex
Varia

@@
 \noindent
-Both of these embody a tradeoff between simplicity, size, and speed.  In this article, we describe both of these strategies and some new variants which offer possible advantages.
+Both of these embody a tradeoff between simplicity, size, and speed.
+Below, we describe both of these strategies and some new variants which offer possible advantages.
+Reversible noun encoding is essential for Nock communication, both between Nock execution layers such as Vere and NockVM, and between the runtime and the host operating system.
 
 \section{Na\"{i}ve Serialization}
+
+A noun may be an atom or a cell, and so we first propose to label nodes by type, with \texttt{0} for atom and \texttt{1} for cell.
+
+\begin{figure}[ht]
+\centering
+\begin{subfigure}
+\centering
+\begin{tikzpicture}[
+  level 1/.style={sibling distance=40mm, line width=1pt},
+  level 2/.style={sibling distance=20mm, line width=1pt},
+  every node/.style={circle,draw, line width=1pt}
+  ]
+  \node {}
+    child { node {\texttt{a}} }
+    child { node {}
+      child { node {\texttt{b}} }
+      child { node {\texttt{c}} }
+    };
+\end{tikzpicture}
+\end{subfigure}
+\begin{subfigure}
+\centering
+\begin{tikzpicture}[
+  level 1/.style={sibling distance=40mm, line width=1pt},
+  level 2/.style={sibling distance=20mm, line width=1pt},
+  every node/.style={square,draw, line width=1pt}
+  ]
+  \node {1}
+    child { node {0} }
+    child { node {1}
+      child { node {0} }
+      child { node {0} }
+    };
+\end{tikzpicture}
+\end{subfigure}
+\caption{The noun \texttt{[a b c]} (the same as \texttt{[a [b c]]}) as a binary tree and its metadata labeling in the simplest scheme.}
+\label{fig:binary-tree}
+\end{figure}
+
+\noindent
+The noun in Figure~\ref{fig:binary-tree} is represented as a binary tree, where each node is labeled with its type (atom or cell) and its value (if applicable).
+To serialize this noun, we can perform a pre-order traversal of the tree, recording the type and value of each node.
+The serialized form would be (in \textsc{lsb} order):
+
+\begin{lstlisting}[style=listingcode]
+c 0 b 0 1 a 0 1
+\end{lstlisting}
+
+\noindent
+The issue, however, is that in decoding this bitstream we do not know where an atom value (such as \texttt{a}) ends.
+After all, the point of serialization is to be able to deserialize the bitstream back into the original noun.
+Thus we need a way to indicate the length of each atom.
+
+Let's augment this scheme by providing length information for each atom.
+One basic idea is to include a string of \texttt{0} bits as the least-significant bits (\textsc{lsb}) of the atom which indicates the length of the atom in bits.
+Since the atom could be zero, we also delimit the length with a \texttt{1} bit.
+(See Table~\ref{tab:atom-encodings} for examples.)
+
+$$
+a = a_{\ell-1} a_{\ell-2} \ldots a_{1} a_{0}
+\xrightarrow[\textrm{encode}]{}
+\bar{a} = a_{\ell-1} a_{\ell-2} \ldots a_{1} a_{0} \; \mathtt{1} \; \mathtt{0}^{n}
+\textrm{,}
+$$
+
+\noindent
+where $\bar{a}$ is the encoded atom $a$ and $a$ has length $\ell$ bits.\footnote{We will use the overbar notation $\bar{a}$ throughout to notionally indicate the encoded form of $a$, regardless of the method used.}
+
+\begin{table}[ht]
+\centering
+\caption{Example atom encodings with length information appended as trailing zeros.}
+\label{tab:atom-encodings}
+\begin{tabular}{ l l r }
+  \textbf{Atom} & \textbf{Length (bits)} & \multicolumn{1}{l}{\textbf{Encoded Form}} \\
+  \hline
+  \texttt{0b0} & 1 & \texttt{0b010} \\
+  \texttt{0b1} & 1 & \texttt{0b110} \\
+  \texttt{0b10} & 2 & \texttt{0b10100} \\
+  \texttt{0b11} & 2 & \texttt{0b11100} \\
+  \texttt{0b101} & 3 & \texttt{0b1011000} \\
+  \texttt{0b111} & 3 & \texttt{0b1111000} \\
+  \texttt{0x70} & 7 & \texttt{0b111000010000000} \\
+  \texttt{0xff} & 8 & \texttt{0b11111111100000000} \\
+\end{tabular}
+\end{table}
+
+For instance, the atom \texttt{0b101010} (decimal \texttt{42}) could be encoded as:
+
+\begin{lstlisting}[style=listingcode]
+:: atom 0x2a
+> `@ub`(gel 0b101010)
+0b101010
+:: cell [0x2a 0x3b]
+> `@ub`(gel [0b101010 0b111011])
+0b1.101010@0.111011
+\end{lstlisting}
 
@@
 ::       0 for end of length; 0 for value
-> `@ub`(jel 0b0)
+> `@ub`(gel 0b0)
 0b10
@@
 ::       0 for end of length; 1 for value
-> `@ub`(jel 0b1)
+> `@ub`(gel 0b1)
 0b1010
@@
 :: [0x0 0x1] = 1 for cell; zero, then one
-> `@ub`(jel [@0b0@ 0b1])
+> `@ub`(gel [@0b0@ 0b1])
 0b1.010@0.010@1
@@
 :: [0x1 0x0] = 1 for cell; one, then zero
-> `@ub`(jel [@0b0@ 0b1])
+> `@ub`(gel [@0b0@ 0b1])
 0b10@1.010@1
@@
 ::       0 for end of length; 2 for value
-> `@ub`(jel 0b10)
+> `@ub`(gel 0b10)
 0b10.0111
@@
 :: 0xff
-> `@ub`(jel 0xff)
+> `@ub`(gel 0xff)
 0b11.1111.1101.1111.1110
@@
 :: [[0x0 0x1] [0x2 0x3]]
-> `@ub`(jel [@0b0@ 0b1])
+> `@ub`(gel [@0b0@ 0b1])
 0b10@1.010@1
@@
 
-Our code implementation for \texttt{++jel} is as follows:
+Our code implementation for \texttt{++gel} is as follows:
 
@@
 !:  |%
-++  jel
-  =jel !:  |=  a=*
+++  gel
+  =gel !:  |=  a=*
   ^-  @
@@
 
-Whatever the pedagogical advantages of \texttt{++jel}, the algorithm has practical flaws:  it is subject to collisions XXX
-and it is verbose.\footnote{Note the claim of \patp{dozreg-toplud}, p.~TODO of this issue, that an operational Arvo instance may have up to $1.66 \times 10^{21}$ nouns, reduced by structural sharing.}  \texttt{++jam} improves the basic strategy by altering the \textsc{rle} algorithm slightly and supporting internal references for noun subtrees that have already been encoded.
+Whatever the pedagogical advantages of \texttt{++gel}, the algorithm has practical flaws:  it is subject to collisions XXX
+and it is verbose.\footnote{Note the claim of \patp{dozreg-toplud}, p.~TODO of this issue, that an operational Arvo instance may have up to $1.66 \times 10^{21}$ nouns, greatly reduced by structural sharing.}  \texttt{++jam} improves the basic strategy by altering the \textsc{rle} algorithm slightly and supporting internal references for noun subtrees that have already been encoded.
 

### 2025-11-02 mss.tex
Post

@@
 \usepackage{ustj}
+\usepackage{cancel}
+\usepackage{pdflscape}
 
@@
 \title{Serializing Nouns}
-\author{N.\ E.\ Davis \patp{lagrev-nocfep}, Brian Klatt \patp{zod},\thanks{Brian Klatt contributed the description of \texttt{++jam}.} \\ \& Sam Parker \patp{zod}\thanks{Sam Parker contributed the description of \texttt{++bulk}.} \\ \affiliation}
+\author{N.\ E.\ Davis \patp{lagrev-nocfep}, Brian Klatt \patp{zod},\thanks{Brian Klatt contributed the description of \texttt{+jam}.} \\ \& Sam Parker \patp{zod}\thanks{Sam Parker contributed the description of \texttt{+bulk}.
+\\ We gratefully acknowledge the contributions of the original architects of \texttt{+jam}/\texttt{+cue} as well.} \\ \affiliation}
 % TODO Brian Klatt credit
@@
   \centering
-\begin{subfigure}
+  % ---------- left tree ----------
+  \begin{subfigure}{\textwidth}
     \centering
@@
         level 2/.style={sibling distance=20mm, line width=1pt},
-  every node/.style={circle,draw, line width=1pt}
+        every node/.style={circle, draw, line width=1pt, minimum size=1.2em}
       ]
       \node {}
-    child { node {\texttt{a}} }
+        child { node {$a$} }
         child { node {}
-      child { node {\texttt{b}} }
-      child { node {\texttt{c}} }
+          child { node {$b$} }
+          child { node {$c$} }
         };
@@
   \end{subfigure}
-\begin{subfigure}
+  \\
+  % ---------- right tree ----------
+  \begin{subfigure}{\textwidth}
     \centering
@@
         level 2/.style={sibling distance=20mm, line width=1pt},
-  every node/.style={square,draw, line width=1pt}
+        every node/.style={rectangle, draw, line width=1pt, minimum size=1.2em}
       ]
-  \node {1}
-    child { node {0} }
-    child { node {1}
-      child { node {0} }
-      child { node {0} }
+      \node {\texttt{1}}
+        child { node {\texttt{0}} }
+        child { node {\texttt{1}}
+          child { node {\texttt{0}} }
+          child { node {\texttt{0}} }
         };
@@
   \end{subfigure}
+
   \caption{The noun \texttt{[a b c]} (the same as \texttt{[a [b c]]}) as a binary tree and its metadata labeling in the simplest scheme.}
@@
 \noindent
-where $\bar{a}$ is the encoded atom $a$ and $a$ has length $\ell$ bits.\footnote{We will use the overbar notation $\bar{a}$ throughout to notionally indicate the encoded form of $a$, regardless of the method used.}
+where $\bar{a}$ is the encoded atom $a$ and $a$ has length $\ell$ bits.\footnote{We will use the overbar notation $\bar{a}$ throughout to notionally indicate the encoded form of $a$, regardless of the method used.}  This is a unary run-length encoding (\textsc{rle}) using zeros.\footnote{Zero is something of a ``special case'':  it has no length and leading zeros are trimmed, so its encoding consists only of the delimiter, \texttt{0b1}.}
 
@@
 \centering
-\caption{Example atom encodings with length information appended as trailing zeros.}
+\caption{Example atom encodings with length information appended as trailing zeros (unnamed encoding).}
 \label{tab:atom-encodings}
 \begin{tabular}{ l l r }
-  \textbf{Atom} & \textbf{Length (bits)} & \multicolumn{1}{l}{\textbf{Encoded Form}} \\
+  \textbf{Atom} $a$ & \textbf{Length (bits)} & \multicolumn{1}{l}{\textbf{Encoded Form} $\bar{a}$} \\
   \hline
-  \texttt{0b0} & 1 & \texttt{0b010} \\
+  \texttt{0b0} & 1 & \texttt{0b1} \\
   \texttt{0b1} & 1 & \texttt{0b110} \\
@@
 
-For instance, the atom \texttt{0b101010} (decimal \texttt{42}) could be encoded as:
+\noindent
+We can make one further optimization:  since the most significant bit (\textsc{msb}) of the atom is always \texttt{1} except in the case of zero (which is \texttt{0b10} with the leading zero trimmed), we can omit it from the encoded form.
+This saves one bit for every non-zero atom.
+Let's call this scheme \texttt{+fat} encoding.
+(See Table~\ref{tab:fat-encodings} for examples.)
 
-\begin{lstlisting}[style=listingcode]
-:: atom 0x2a
-> `@ub`(gel 0b101010)
-0b101010
-:: cell [0x2a 0x3b]
-> `@ub`(gel [0b101010 0b111011])
-0b1.101010@0.111011
-\end{lstlisting}
+\paragraph*{\texttt{+fat} Serialization Algorithm}
 
-The simplest way to encode a noun as a binary tree in an atom (byte array) without compression is to utilize bits to mark atom (\texttt{0}) vs.
-cell (\texttt{1}), the length of the atom in unary of \texttt{1}s terminated by \texttt{0}, and the actual value.
-(Since the leading zeros are stripped, they may need to be added back in.)  This is in least-significant \emph{byte} (\textsc{lsb}) order, so you should read the written atom starting from the \emph{right} in binary.
+\begin{itemize}
+\item  \textbf{Encode}.
+Write\footnote{We always say this from the \textsc{lsb}, so from the right as written on the page.} $\ell$ bits of \texttt{0}, where $\ell$ is the length of $a$ in unary, followed by \texttt{1}.
+The next $\ell$ bits are the atom $a$ in binary without the leading \texttt{1}.
+\item  \textbf{Decode}.
+Read the unary length $\ell$ by counting \texttt{0} bits until the first \texttt{1} bit.
+Read the next $\ell$ bits to get the atom $a$ without the leading \texttt{1}.
+Prepend a \texttt{1} bit to $a$ unless $\ell$ is zero, in which case the atom is zero.
+\end{itemize}
 
-Thus the atom \texttt{0b0} would simply be encoded as:
+\begin{table}[ht]
+\centering
+\caption{Example atom encodings (\texttt{+fat} encoding).}
+\label{tab:fat-encodings}
+\begin{tabular}{ l l r }
+  \textbf{Atom} $a$ & \textbf{Length (bits)} & \multicolumn{1}{l}{\textbf{Encoded Form} $\bar{a}$} \\
+  \hline
+  \texttt{0b0} & 1 & \texttt{0b1} \\
+  \texttt{0b1} & 1 & \texttt{0b10} \\
+  \texttt{0b10} & 2 & \texttt{0b100} \\
+  \texttt{0b11} & 2 & \texttt{0b1100} \\
+  \texttt{0b101} & 3 & \texttt{0b11000} \\
+  \texttt{0b111} & 3 & \texttt{0b111000} \\
+  \texttt{0x70} & 7 & \texttt{0b11000010000000} \\
+  \texttt{0xff} & 8 & \texttt{0b1111111100000000} \\
+\end{tabular}
+\end{table}
 
-\begin{lstlisting}[style=listingcode]
-:: 0x0 = 0 for atom; 1 for length (special case);
-::       0 for end of length; 0 for value
-> `@ub`(gel 0b0)
-0b10
+The primary problem with the \texttt{+fat} scheme is that the encoding of an atom is twice the bit length of the original atom.
+The next optimization is to use the bit length of the bit length, which is logarithmic in the bit length.
+
+$$
+a =
+\underbrace{a_{\ell-1} a_{\ell-2} \ldots a_{1} a_{0}}
+_{\textstyle \mathrlap{\kern-0.25em\ell =
+\underbrace{\ell_{h-1} \ell_{h-2} \ldots \ell_{1} \ell_{0}}
+_{\textstyle h}}}
+$$
+
+\begin{align*}
+a = & a_{\ell-1} a_{\ell-2} \ldots a_{1} a_{0}
+\xrightarrow[\textrm{\texttt{+mat} encode}]{}
+\\
+\bar{a} = &
+\underbrace{a_{\ell-1} a_{\ell-2} \ldots a_{1} a_{0}}
+_{\ell}
+\underbrace{\ell_{h-2} \ell_{h-1} \ldots \ell_{1} \ell_{0}}
+_{h-1}
+\mathtt{1}
+\underbrace{\mathtt{0} \ldots \mathtt{0}}
+_{h}
+\textrm{,}
+\end{align*}
+
+\noindent
+where $\ell$ is the bit length of $a$ and $h$ is the ``hyper bit length'' of $a$.
+This yields a bit length encoding of $1+2\log_2(\ell)$ bits rather than $2\ell$ bits.
+This scheme is called \lstinline[style=inlinecode]{+mat} encoding, and it is currently in use as an internal auxiliary format for other serializations.
+
+\paragraph*{\texttt{+mat} Serialization Algorithm}
+
+\begin{itemize}
+\item  \textbf{Encode}.
+Write the number $h$ of \texttt{0} bits in unary, followed a leading \texttt{1} and then $h-1$ bits for the low bits of $\ell$.
+The next $\ell$ bits are the atom $a$ in binary.
+\item  \textbf{Decode}.
+Read the unary length of the length $h$ by counting \texttt{0} bits (from the \textsc{lsb}/right) until the first \texttt{1} bit.
+Read the next $h-1$ bits to get $\ell-1$, the bit length of the atom.
+Finally, read the next $\ell$ bits (adding the leading \texttt{1}) to get the atom $a$.
+\end{itemize}
+
+Some worked examples of \texttt{+mat} encoding are included in Table~\ref{tab:mat-encodings}.
+
+\begin{landscape}
+\begin{table}[ht]
+\centering
+\caption{Example atom encodings (\texttt{+mat} encoding).  The encoded form is the same result as \texttt{q:(mat a)} in Hoon.}
+\label{tab:mat-encodings}
+\begin{tabular}{ r c c r }
+  \multicolumn{1}{l}{\textbf{Atom}} $a$ & \textbf{Length (bits)} & \textbf{Mathematics} & \multicolumn{1}{l}{\textbf{Encoded Form} $\bar{a}$} \\
+  \hline
+  \texttt{0b0} & $\ell = 1$ & $\underbrace{\cancel{{a_{\ell-1}}}\ldots\cancel{a_{0}}}_{\ell=0}\mathtt{1}\underbrace{\cancel{\mathtt{0}}\ldots\cancel{\mathtt{0}}}_{h=0}$ & \texttt{0b1} \\
+  \texttt{0b1} & $\ell = 1$ & $\underbrace{\mathtt{1}}_{\ell=1}\underbrace{\cancel{{\ell_{h-2}}}\ldots\cancel{\ell_{0}}}_{h-1=0}\mathtt{1}\underbrace{\mathtt{0}}_{h=1}$ & \texttt{0b110} \\
+  \texttt{0b10} & $\ell = 2$ & $\underbrace{\mathtt{10}}_{a=2}\underbrace{\mathtt{0}}_{\textrm{low bits of }\ell}\mathtt{1}\underbrace{\mathtt{00}}_{h=2}$ & \texttt{0b100100} \\
+  \texttt{0b11} & $\ell = 2$ & $\underbrace{\mathtt{11}}_{a=3}\underbrace{\mathtt{0}}_{\textrm{low bits of }\ell}\mathtt{1}\underbrace{\mathtt{00}}_{h=2}$ & \texttt{0b110100} \\
+  \texttt{0b100} & $\ell = 3$ & $\underbrace{\mathtt{100}}_{a=4}\underbrace{\mathtt{1}}_{\textrm{low bits of }\ell}\mathtt{1}\underbrace{\mathtt{00}}_{h=2}$ & \texttt{0b1001100} \\
+  \texttt{0b101} & $\ell = 3$ & $\underbrace{\mathtt{101}}_{a=5}\underbrace{\mathtt{1}}_{\textrm{low bits of }\ell}\mathtt{1}\underbrace{\mathtt{00}}_{h=2}$ & \texttt{0b1011100} \\
+  \texttt{0b1111} & $\ell = 4$ & $\underbrace{\mathtt{1111}}_{a=15}\underbrace{\mathtt{00}}_{\textrm{low bits of }\ell}\mathtt{1}\underbrace{\mathtt{000}}_{h=3}$ & \texttt{0b1111001000} \\
+  \texttt{0x70} & $\ell = 7$ & $\underbrace{\mathtt{1110000}}_{a=112}\underbrace{\mathtt{11}}_{\textrm{low bits of }\ell}\mathtt{1}\underbrace{\mathtt{000}}_{h=3}$ & \texttt{0b1110000111000} \\
+\end{tabular}
+\end{table}
+\end{landscape}
+
+With \texttt{+mat} encoding in hand, we can produce a proper serialization scheme for nouns by combining the type bit with the encoded atom or recursively encoding cells.
+We call this scheme \texttt{+gel} encoding.
+
+As before, we use \texttt{0} for atom and \texttt{1} for cell.
+If the noun is an atom, we follow the type bit with the \texttt{+mat} encoding of the atom.
+If the noun is a cell, we follow the type bit with the \texttt{+gel} encoding of the head and then the \texttt{+gel} encoding of the tail.
+
+Consider the noun \texttt{[a [b c]]}, represented as a binary tree in Figure~\ref{fig:binary-tree-example}.  Reading across in a depth-first traversal, we produce an encoding order:
+
+\begin{lstlisting}[style=listingblock]
+(mat 1)(0)(mat 0)(0)(1)(mat 4)(0)(1)
 \end{lstlisting}
@@
 \noindent
-interpreted as (rightmost, LSB) \texttt{0} for atom, run length of \texttt{1}, and value \texttt{0} (stripped).  I.e., \texttt{0b010} or
+We already have the \texttt{+mat} encodings from Table~\ref{tab:mat-encodings}; substituting these in yields:
 
-{
-\centering
-$\leftarrow$ \\
-\bit{0} \bit{1} \bit{0}
-}
-
-Other values include:
-
-\begin{lstlisting}[style=listingcode]
-:: 0x1 = 0 for atom; 1 for length;
-::       0 for end of length; 1 for value
-> `@ub`(gel 0b1)
-0b1010
-
-:: [0x0 0x1] = 1 for cell; zero, then one
-> `@ub`(gel [@0b0@ 0b1])
-0b1.010@0.010@1
-
-:: [0x1 0x0] = 1 for cell; one, then zero
-> `@ub`(gel [@0b0@ 0b1])
-0b10@1.010@1
-
-:: 0x2 = 0 for atom; 2 for length;
-::       0 for end of length; 2 for value
-> `@ub`(gel 0b10)
-0b10.0111
-
-:: 0xff
-> `@ub`(gel 0xff)
-0b11.1111.1101.1111.1110
-
-:: [[0x0 0x1] [0x2 0x3]]
-> `@ub`(gel [@0b0@ 0b1])
-0b10@1.010@1
+\begin{lstlisting}[style=listingblock]
+[110](0)[1](0)(1)[1001100](0)(1)
+0b110.0.1.0.1.1001100.0.1
+0b1100101100110001
 \end{lstlisting}
 
-Our code implementation for \texttt{++gel} is as follows:
+\begin{figure}[ht]
+  \centering
+  % ---------- left tree ----------
+  \begin{subfigure}{\textwidth}
+    \centering
+    \begin{tikzpicture}[
+        level 1/.style={sibling distance=40mm, line width=1pt},
+        level 2/.style={sibling distance=20mm, line width=1pt},
+        every node/.style={circle, draw, line width=1pt, minimum size=1.2em}
+      ]
+      \node {}
+        child { node {4} }
+        child { node {}
+          child { node {0} }
+          child { node {1} }
+        };
+    \end{tikzpicture}
+  \end{subfigure}
+  \\
+  % ---------- right tree ----------
+  \begin{subfigure}{\textwidth}
+    \centering
+    \begin{tikzpicture}[
+        level 1/.style={sibling distance=40mm, line width=1pt},
+        level 2/.style={sibling distance=20mm, line width=1pt},
+        every node/.style={rectangle, draw, line width=1pt, minimum size=1.2em}
+      ]
+      \node {\texttt{1}}
+        child { node {\texttt{0}} }
+        child { node {\texttt{1}}
+          child { node {\texttt{0}} }
+          child { node {\texttt{0}} }
+        };
+    \end{tikzpicture}
+  \end{subfigure}
+
+  \caption{The noun \texttt{[4 0 1]} as a binary tree and its metadata labeling in the simplest scheme.  Compare Figure~\ref{fig:binary-tree}.}
+  \label{fig:binary-tree-example}
+\end{figure}
+
+\paragraph*{\texttt{+gel} Serialization Algorithm}
+\begin{itemize}
+\item  \textbf{Encode}.
+  \begin{itemize}
+    \item  If the noun is an atom, write \texttt{0} followed by the \texttt{+mat} encoding of the atom.
+    \item  If the noun is a cell, write \texttt{1} followed by the \texttt{+gel} encoding of the head and then the \texttt{+gel} encoding of the tail.
+  \end{itemize}
+\item  \textbf{Decode}.  Read the first bit.
+  \begin{itemize}
+    \item  If it is \texttt{0}, read the next bits as a \texttt{+mat} encoded atom.
+    \item  If it is \texttt{1}, recursively read the next bits as the head and then the tail.
+  \end{itemize}
+\end{itemize}
+
+\texttt{+gel} encoding is straightforward to read and write even by hand.  Worked examples are included in Table~\ref{tab:gel-encodings}.
+
+\begin{table}[ht]
+\centering
+\caption{Example noun encodings (\texttt{+gel} encoding).  \texttt{.} dot delimits bit fields for readability.}
+\label{tab:gel-encodings}
+\begin{tabular}{ r r }
+  \multicolumn{1}{l}{\textbf{Noun}} & \multicolumn{1}{l}{\textbf{Encoded Form}} \\
+  \hline
+  \texttt{[0 0]} & \texttt{0b1.0.1.0.1} \\
+  \texttt{[1 0]} & \texttt{0b110.0.1.0.1} \\
+  \texttt{[2 1 0]} & \texttt{0b110.0.1.0.1.100100.0.1} \\
+  \texttt{[[2 3] 1 0]} & \texttt{0b110.0.1.0.1.110100.0.100100.0.1.1} \\
+\end{tabular}
+\end{table}
+
+TODO opposite of gel as goo
+
+Our code implementation for \texttt{+gel} is as follows:
 
@@
 
-Note that \texttt{l} is not the length of an atom in unary, but the length of the encoded noun in binary.
-
 \section{Practical Serialization}
 
-Whatever the pedagogical advantages of \texttt{++gel}, the algorithm has practical flaws:  it is subject to collisions XXX
-and it is verbose.\footnote{Note the claim of \patp{dozreg-toplud}, p.~TODO of this issue, that an operational Arvo instance may have up to $1.66 \times 10^{21}$ nouns, greatly reduced by structural sharing.}  \texttt{++jam} improves the basic strategy by altering the \textsc{rle} algorithm slightly and supporting internal references for noun subtrees that have already been encoded.
+Whatever the pedagogical advantages of \texttt{+gel}, the algorithm is still relatively verbose.
+Nock-based operating functions have preferred in practice to use \texttt{+jam} encoding, which is a more compact run-length encoding with subtree deduplication.\footnote{Note the calculation of \patp{dozreg-toplud}, p.~TODO of this issue, that an operational Arvo instance may have up to $1.66 \times 10^{21}$ nouns, greatly reduced by structural sharing.}  \texttt{+jam} improves the basic strategy by altering the \textsc{rle} algorithm slightly and supporting internal references for noun subtrees that have already been encoded.
 
-\texttt{++jam} converts a noun into a buffer and deduplicates repeated subtrees.
+\texttt{+jam} converts a noun into a buffer and deduplicates repeated subtrees.
 It walks subtrees and encodes each in a way that allows for efficient storage and retrieval, while also permitting references to previously encoded values.
+Basically, it takes \texttt{+gel} and adds a ``caching mode'' flag; instead of repeating subtrees, encode the index in the full jammed noun to where the current node's jammed subnoun was first written.
+Thus the code changes from \texttt{0} for atom and \texttt{1} for cell to:
 
-- get Bryan notes
+\begin{itemize}
+  \item  \texttt{0} for atom.
+  \item  \texttt{01} for cell.
+  \item  \texttt{11} for cached node.
+\end{itemize}
 
-special-cases 0
+\begin{figure}[ht]
+  \centering
+  % ---------- left tree ----------
+  \begin{subfigure}{\textwidth}
+    \centering
+    \begin{tikzpicture}[
+        level 1/.style={sibling distance=40mm, line width=1pt},
+        level 2/.style={sibling distance=20mm, line width=1pt},
+        every node/.style={circle, draw, line width=1pt, minimum size=1.2em}
+      ]
+      \node {}
+        child { node {}
+          child { node {0} }
+          child { node {0} }
+        }
+        child { node {}
+          child { node {0} }
+          child { node {0} }
+        };
+    \end{tikzpicture}
+  \end{subfigure}
+  \\
+  % ---------- right tree ----------
+  \begin{subfigure}{\textwidth}
+    \centering
+    \begin{tikzpicture}[
+        level 1/.style={sibling distance=40mm, line width=1pt},
+        level 2/.style={sibling distance=20mm, line width=1pt},
+        every node/.style={rectangle, draw, line width=1pt, minimum size=1.2em}
+      ]
+      \node {\underline{\texttt{01}}}
+        child { node {\texttt{01}}
+          child { node {\underline{\texttt{0}}} }
+          child { node {\texttt{0}} }
+        }
+        child { node {\texttt{11}}
+          child { node {} }
+          child { node {} }
+        };
+    \end{tikzpicture}
+  \end{subfigure}
 
-The new \textsc{rle} calculation is to post the number plus one in binary rather than unary (e.g., for \texttt{} the length would be \texttt{0b010}).
+  \caption{The noun \texttt{[[0 0] 0 0]} as a binary tree and its metadata labeling for the \texttt{+jam} scheme.}
+  \label{fig:binary-tree-jam}
+\end{figure}
 
-(Since there is a value less significant than the length, the leading zero is not lost but serves as a divider.)
+Following the metadata labels given in Figure~\ref{fig:binary-tree-jam}, the encoding order for \texttt{[[0 0] 0 0]} is:
+
+\begin{lstlisting}[style=listingblock]
+(mat 2)(11)(mat 0)(0)(mat 0)((@$\underline{\mathtt{0}}$@))(01)((@$\underline{\mathtt{01}}$@))
+0b100100.11.1.0.1.(@$\underline{\mathtt{0}}$@).01.(@$\underline{\mathtt{01}}$@)
+0b1001001110100101
+\end{lstlisting}
+
+\noindent
+The cached entry \texttt{(mat 2)} indicates that the cached subtree is a copy of the subtree located at tree address 2 in the overall jammed noun.
+
+However, this bit of cleverness isn't the end:  \texttt{+jam} also decides \emph{when} to cache subtrees based on the size of the nouns involved.  For instance, \texttt{[3 3 3]} encodes as:
+
+\begin{lstlisting}[style=listingblock]
+(mat 3)(0)(mat 3)(0)(01)(mat 3)(0)(01)
+0b110100.0.110100.0.01.110100.0.01
+0b1101000110100001110100001
+\end{lstlisting}
+
+\noindent
+wherein no caching is used.
+When 3 is first encountered at index 2, 2 and 3 have the same bit lengths; since \texttt{(mat 2)} and \texttt{(mat 3)} are the same length, caching would not save any space and would add decoding overhead due to dereferencing.
+
+In contrast, \texttt{[4 4 4]} encodes as:
+
+\begin{lstlisting}[style=listingblock]
+(mat 2)(11)(mat 2)(11)(01)(1001100)(0)(01)
+0b100100.11.100100.11.01.1001100.0.01
+0b1001001110010011011001100001
+\end{lstlisting}
+
+\noindent
+where caching is used for the second and third occurrences of 4.  Here, 4 has bit length 3 which is greater than the bit length of 2 which is 2, and caching saves space overall in the jammed noun.
+
+\paragraph*{\texttt{+jam} Serialization Algorithm}
+
+\begin{itemize}
+\item  \textbf{Encode}.
+  \begin{itemize}
+    \item  If the noun is an atom, write \texttt{0} followed by the \texttt{+mat} encoding of the atom.
+    \item  If the noun is a cell, check if it has been previously encoded.
+      \begin{itemize}
+        \item  If it has, write \texttt{11} followed by the \texttt{+mat} encoding of the index where it was first encoded.
+        \item  If it has not, write \texttt{01} followed by the \texttt{+jam} encoding of the head and then the \texttt{+jam} encoding of the tail.  Store the current index for future reference.
+      \end{itemize}
+  \end{itemize}
+\item  \textbf{Decode}.  Read the first bit.
+  \begin{itemize}
+    \item  If it is \texttt{0}, read the next bits as a \texttt{+mat} encoded atom.
+    \item  If it is \texttt{1}, read the next bit.
+      \begin{itemize}
+        \item  If it is \texttt{0}, recursively read the next bits as the head and then the tail.
+        \item  If it is \texttt{1}, read the next bits as a \texttt{+mat} encoded index and retrieve the cached subtree.
+      \end{itemize}
+  \end{itemize}
+\end{itemize}
+
+Here is the reference implementation of \texttt{+jam} in Hoon:
 
@@
 
-A Python example of \texttt{++jam} is included in Appendix~A.
+A Python implementation of \texttt{+jam} is included in Appendix~A.
+
+Here is the reference implementation of \texttt{+cue} in Hoon:
+
+\begin{lstlisting}[style=listingcode]
+TODO
+\end{lstlisting}
+
+Likewise, a Python implementation of \texttt{+cue} is included in Appendix~A.
 
@@
 \sloppy
-“Newt” encoding is a runtime-oriented extension of \texttt{++jam}\-based noun serialization which adds a short identifying header in case of future changes to the serialization format.
-A \mbox{version} number (currently a single bit) precedes a \textsc{rle} serialization length followed by the \texttt{++jam} serialization of the noun.
+“Newt” encoding is a runtime-oriented extension of \texttt{+jam}\-based noun serialization which adds a short identifying header in case of future changes to the runtime's serialization format.
+A \mbox{version} number (currently a single bit) precedes a \textsc{rle} serialization length followed by the \texttt{+jam} serialization of the noun.
 The version number is currently \texttt{0b0}.
@@
 \noindent
-where \texttt{V} is the version number, \texttt{L} is the total length of the noun in bytes, and \texttt{J} is the \texttt{++jam} serialization of the noun.
+where \texttt{V} is the version number, \texttt{L} is the total length of the noun in bytes, and \texttt{J} is the \texttt{+jam} serialization of the noun.
 
@@
 
-A directed graph encoding has been independently proposed twice, once by Tlon in the original Hoon codebase as a \lstinline[style=inlinecode]{+$silo} encoding and once by Sam Parker as the \lstinline[style=inlinecode]{++bulk} encoding.
+A directed graph encoding has been independently proposed twice, once by Tlon in the original Hoon codebase as a \lstinline[style=inlinecode]{+$silo} encoding and once by Sam Parker as the \lstinline[style=inlinecode]{+bulk} encoding.
 
@@
 
-One of the advantages of \texttt{++jam} is its compactness.
-However, this comes at the cost of speed, since bit-level operations are required to \texttt{++cue} the noun back from its serialized form.
+One of the advantages of \texttt{+jam} is its compactness.
+However, this comes at the cost of speed, since bit-level operations are required to \texttt{+cue} the noun back from its serialized form.
 If a slightly larger size is acceptable, a byte-aligned serialization could facilitate certain kinds of external inspection without requiring deserialization.
-(For instance, a byte-aligned head tag could be read for a rapid decision without needing to \texttt{++cue} the entire noun.)
+(For instance, a byte-aligned head tag could be read for a rapid decision without needing to \texttt{+cue} the entire noun.)
 
-We propose a strategy to modify \texttt{++jam} to align to bytes by padding the length of entries to the nearest byte boundary and marking the distance with a clever binary scheme rather than simply unary.
-This approach, called \texttt{++honey}, aims to balance compactness and speed for certain use cases while retaining a large degree of conceptual backwards compatibility.
+We propose a strategy to modify \texttt{+jam} to align to bytes by padding the length of entries to the nearest byte boundary and marking the distance with a clever binary scheme rather than simply unary.
+This approach, called \texttt{+honey}, aims to balance compactness and speed for certain use cases while retaining a large degree of conceptual backwards compatibility.
 (The change in byte alignment of course breaks strict compatibility.)
@@
 
-\section*{Appendix A:  Python \texttt{++jam}/\texttt{++cue}}
+\section*{Appendix A:  Python \texttt{+jam}/\texttt{+cue}}
 
-The following is a simple Python implementation of \texttt{++jam} serialization drawn from Urbit’s auxiliary \texttt{pynoun} library.
+The following is a simple Python implementation of \texttt{+jam} serialization drawn from Urbit’s auxiliary \texttt{pynoun} library.
 

### 2025-11-03 mss.tex
Add graph representation.

@@
 
-\newcommand{\authorname}{N. E. Davis, Brian Klatt, \& Sam Parker}
-\newcommand{\authorpatp}{\patp{lagrev-nocfep}, \patp{zod}, \& \patp{zod}}
-\newcommand{\affiliation}{Zorp Corp, Zorp Corp, \& —}
+\newcommand{\authorname}{N. E. Davis \patp{lagrev-nocfep},* Brian Klatt \patp{ticlys-monlun},* \& Sam Parker \patp{borlug-walryd}†}
+% \newcommand{\authorpatp}{, , \& \patp{borlug-walryd}}
+\newcommand{\affiliation}{* Zorp Corp \\ † —}
 
@@
 \title{Serializing Nouns}
-\author{N.\ E.\ Davis \patp{lagrev-nocfep}, Brian Klatt \patp{zod},\thanks{Brian Klatt contributed the description of \texttt{+jam}.} \\ \& Sam Parker \patp{zod}\thanks{Sam Parker contributed the description of \texttt{+bulk}.
-\\ We gratefully acknowledge the contributions of the original architects of \texttt{+jam}/\texttt{+cue} as well.} \\ \affiliation}
-% TODO Brian Klatt credit
+\author{N. E. Davis \patp{lagrev-nocfep},* \\ Brian Klatt \patp{ticlys-monlun},* \\ \& Sam Parker \patp{borlug-walryd}† \\ \affiliation}
 \date{}
@@
 Noun serialization is commonly used for Nock communication, both between instances like Urbit ships and with the runtime and the Unix host operating system.
-This article describes and compares the two principal conventions for representing nouns in slightly compressed form as byte arrays, as well as introduces variant encodings for educational purposes.
+This article describes and compares the two principal conventions for representing nouns in slightly compressed form as byte arrays, as well as introduces variant encodings for educational purposes.\footnote{Brian Klatt contributed the description of \texttt{+jam}.
+Sam Parker contributed the description of \texttt{+bulk}.
+We gratefully acknowledge the contributions of the original architects of \texttt{+jam}/\texttt{+cue} as well.}
 \end{abstract}
@@
 
-A directed graph encoding has been independently proposed twice, once by Tlon in the original Hoon codebase as a \lstinline[style=inlinecode]{+$silo} encoding and once by Sam Parker as the \lstinline[style=inlinecode]{+bulk} encoding.
+A directed graph encoding has been independently proposed twice, once by Tlon in the original Hoon codebase as a \lstinline[style=inlinecode]{+$silo TODO} encoding and once by Sam Parker as the \lstinline[style=inlinecode]{+bulk} encoding.
+
+\begin{figure}[htp]
+  \centering
+    \centering
+    \begin{subfigure}{\textwidth}
+      \centering
+      \begin{tikzpicture}[
+          level 1/.style={sibling distance=40mm, line width=1pt},
+          level 2/.style={sibling distance=20mm, line width=1pt},
+          level 3/.style={sibling distance=10mm, line width=1pt},
+          level 4/.style={sibling distance=10mm, line width=1pt},
+          every node/.style={circle, draw, line width=1pt, minimum size=1.2em},
+          every node/.style={circle, draw, minimum size=0.5cm, thick},
+          >=stealth
+        ]
+        % 1
+        \node[very thick] {}
+          child[edge from parent/.style={draw, ->, very thick}] { 
+            %2
+            node[double, thick] {}
+            child[edge from parent/.style={draw, ->, double, thick}] { 
+              %4
+              node[dashed, thick] {}
+              child[edge from parent/.style={draw, ->, dashed, thick}] { 
+                %8
+                node {} }
+              child[edge from parent/.style={draw, ->, dashed, thick}] { 
+                %9
+                node[dotted, thick] {}
+                child[edge from parent/.style={draw, ->, dotted, thick}] { 
+                  %18
+                  node[solid, thick] {} }
+                child[edge from parent/.style={draw, ->, dotted, thick}] { 
+                  %19
+                  node[solid, thick] {} }
+              }
+            }
+            child[edge from parent/.style={draw, ->, double, thick}] { 
+              %5
+              node[dotted, thick] {}
+              child[edge from parent/.style={draw, ->, dotted, thick}] { 
+                %10
+                node {} }
+              child[edge from parent/.style={draw, ->, dotted, thick}] { 
+                %11
+                node {} }
+            }
+          }
+          child[edge from parent/.style={draw, ->, very thick}] { 
+            %3
+            node[dashed, thick] {}
+            child[edge from parent/.style={draw, ->, dashed, thick}] { 
+              %6
+              node {} }
+            child[edge from parent/.style={draw, ->, dashed, thick}] { 
+              %7
+              node[dotted, thick] {}
+              child[edge from parent/.style={draw, ->, dotted, thick}] { 
+                %14
+                node[solid, thick] {} }
+              child[edge from parent/.style={draw, ->, dotted, thick}] { 
+                %15
+                node[solid, thick] {} }
+            }
+          };
+      \end{tikzpicture}
+    \end{subfigure}
+    \\
+    \begin{subfigure}{\textwidth}
+      \centering
+      \begin{tikzpicture}[
+        node distance=3cm,
+        every node/.style={circle, draw, minimum size=0.5cm, thick},
+        >=stealth
+      ]
+        % Define the four nodes in a square
+        \node[vertex, very thick] (Root) at (1.5,4.5) {};
+        \node[vertex, double, thick] (A) at (0,3) {};
+        \node[vertex, dashed, thick] (B) at (3,3) {};
+        \node[vertex, dotted, thick] (C) at (0,0) {};
+        \node[vertex, thick] (D) at (3,0) {};
+
+        \draw[->, very thick] (Root) -- node[above, draw=none] {L} (A);
+        \draw[->, very thick] (Root) -- node[above, draw=none] {R} (B);
+
+        \draw[->, double, thick] (A) -- node[above, draw=none] {L} (B);
+        \draw[->, double, thick] (A) -- node[left, draw=none] {R} (C);
+
+        \draw[->, dashed, thick] (B) -- node[right, draw=none] {R} (C);
+        \draw[->, dashed, thick] (B) -- node[left, draw=none] {L} (D);
+
+        \draw[->, dotted, thick, bend left=20] (C) to node[above, draw=none] {L} (D);
+        \draw[->, dotted, thick, bend right=20] (C) to node[below, draw=none] {R} (D);
+      \end{tikzpicture}
+    \end{subfigure}
+  \caption{The noun \texttt{[[a [a a]] [a a]] [a [a a]]} as a binary tree and its metadata labeling in a directed graph representation.}
+  \label{fig:binary-tree-dcg}
+\end{figure}
 
@@
 \section*{Appendix A:  Python \texttt{+jam}/\texttt{+cue}}
+\subsection*{\patp{fodwyt-ragful}}
 
-The following is a simple Python implementation of \texttt{+jam} serialization drawn from Urbit’s auxiliary \texttt{pynoun} library.
+The following is a simple Python implementation of \texttt{+jam} serialization drawn from Urbit’s auxiliary \texttt{pynoun} library.\footnote{The original \texttt{pynoun} library was authored by Paul Driver \patp{fodwyt-ragful}.}
 

### 2025-11-05 mss.tex
Realign +fat

--- 2025-11-02 mss.tex
@@
 
-\newcommand{\authorname}{N. E. Davis, Brian Klatt, \& Sam Parker}
-\newcommand{\authorpatp}{\patp{lagrev-nocfep}, \patp{zod}, \& \patp{zod}}
-\newcommand{\affiliation}{Zorp Corp, Zorp Corp, \& —}
+\newcommand{\authorname}{Brian Klatt \patp{ticlys-monlun}, N. E. Davis \patp{lagrev-nocfep}}
+\newcommand{\affiliation}{Zorp Corp}
 
@@
 \fancyhead[LE]{{\urbitfont Urbit Systems Technical Journal}}
-\fancyhead[RO]{Serializing Nouns}
+\fancyhead[RO]{Noun Serialization}
 \fancyfoot[LE,RO]{\thepage}
@@
 %%MANUSCRIPT
-\title{Serializing Nouns}
-\author{N.\ E.\ Davis \patp{lagrev-nocfep}, Brian Klatt \patp{zod},\thanks{Brian Klatt contributed the description of \texttt{+jam}.} \\ \& Sam Parker \patp{zod}\thanks{Sam Parker contributed the description of \texttt{+bulk}.
-\\ We gratefully acknowledge the contributions of the original architects of \texttt{+jam}/\texttt{+cue} as well.} \\ \affiliation}
-% TODO Brian Klatt credit
+\title{Noun Serialization}
+\author{Brian Klatt \patp{ticlys-monlun}, \\ N. E. Davis \patp{lagrev-nocfep}* \\ \affiliation}
 \date{}
@@
 Noun serialization is commonly used for Nock communication, both between instances like Urbit ships and with the runtime and the Unix host operating system.
-This article describes and compares the two principal conventions for representing nouns in slightly compressed form as byte arrays, as well as introduces variant encodings for educational purposes.
+This article describes the principal convention for representing a noun in compressed form as a byte array, as well as introduces variant encodings for educational purposes.
 \end{abstract}
+
+\renewcommand{\thefootnote}{\fnsymbol{footnote}}
+\setcounter{footnote}{0}
+\footnotetext[1]{\sloppy We gratefully acknowledge the original architect of \texttt{+jam} and \texttt{+cue}, \mbox{Curtis} Yarvin \patp{sorreg-namtyv}.
+The original function, in \href{https://github.com/cgyarvin/urbit/blame/55f591594080d276fc446ceafa4afff092561613/watt/jam.watt}{\texttt{jam.watt}}, dates to \textasciitilde 2011.6.11.}
+\renewcommand{\thefootnote}{\arabic{footnote}}
+\setcounter{footnote}{0}
 
@@
 Unlike Gödel numbering, which classically serially encodes the symbols of mathematical statements, noun serialization encodes a binary tree structure.\footnote{This also echoes the way that S-expressions are encoded in Lisp.}  Because a noun may be any atom---and atoms cannot have leading zeroes---both structure and value need to be unambiguously encoded and cannot be simply delimited (as by a \texttt{0} bit or similar).
-There are two basic strategies to encode a noun as an atom:
-
-\begin{enumerate}
-  \item  Run-length serialization, with or without references.
-  \item  Directed graph serialization, depending on a reentrant graph encoding.
-\end{enumerate}
-
-\noindent
-Both of these embody a tradeoff between simplicity, size, and speed.
-Below, we describe both of these strategies and some new variants which offer possible advantages.
-Reversible noun encoding is essential for Nock communication, both between Nock execution layers such as Vere and NockVM, and between the runtime and the host operating system.
+Run-length serialization, with or without references, forms the basis of practical strategies to encode a noun as an atom.
+This embodies a tradeoff between simplicity, size, and speed.
+We describe this strategy and some pedagogical variants which afford insight to the process.
+Likewise, deserialization is the inverse operation of converting a serialized atom back into the original noun, and must be unambiguously reversible.
+Noun serialization is essential for Nock communication, both between Nock execution layers such as Vere and NockVM, and between the runtime and the host operating system.
 
@@
 
+\subsection{Atom Length Encoding}
+
 Let's augment this scheme by providing length information for each atom.
@@
   \texttt{0xff} & 8 & \texttt{0b11111111100000000} \\
+  \texttt{0x1234} & 15 & \texttt{0b100001100100001} \\
+  & & \texttt{↳1000000000000000} \\
 \end{tabular}
@@
 \noindent
-We can make one further optimization:  since the most significant bit (\textsc{msb}) of the atom is always \texttt{1} except in the case of zero (which is \texttt{0b10} with the leading zero trimmed), we can omit it from the encoded form.
+We can make one further optimization:  since the most significant bit (\textsc{msb}) of the atom is always \texttt{1} except in the case of zero (which is \texttt{0b1} with the leading zero trimmed), we can omit it from the encoded form.
 This saves one bit for every non-zero atom.
@@
 \centering
-\caption{Example atom encodings (\texttt{+fat} encoding).}
+\caption{Example atom encodings (\texttt{+fat} encoding).  Elided digits are shown in light gray for clarity.}
 \label{tab:fat-encodings}
@@
   \texttt{0b0} & 1 & \texttt{0b1} \\
-  \texttt{0b1} & 1 & \texttt{0b10} \\
-  \texttt{0b10} & 2 & \texttt{0b100} \\
-  \texttt{0b11} & 2 & \texttt{0b1100} \\
-  \texttt{0b101} & 3 & \texttt{0b11000} \\
-  \texttt{0b111} & 3 & \texttt{0b111000} \\
-  \texttt{0x70} & 7 & \texttt{0b11000010000000} \\
-  \texttt{0xff} & 8 & \texttt{0b1111111100000000} \\
+  \texttt{0b1} & 1 & \texttt{0b\textcolor{lightgray}{1}10} \\
+  \texttt{0b10} & 2 & \texttt{0b\textcolor{lightgray}{10}100} \\
+  \texttt{0b11} & 2 & \texttt{0b\textcolor{lightgray}{1}1100} \\
+  \texttt{0b101} & 3 & \texttt{0b\textcolor{lightgray}{10}11000} \\
+  \texttt{0b111} & 3 & \texttt{0b\textcolor{lightgray}{1}111000} \\
+  \texttt{0x70} & 7 & \texttt{0b\textcolor{lightgray}{1}11000010000000} \\
+  \texttt{0xff} & 8 & \texttt{0b\textcolor{lightgray}{1}1111111100000000} \\
+  \texttt{0x1234} & 15 & \texttt{0b\textcolor{lightgray}{10000}1100100001} \\
+  & & \texttt{↳1000000000000000} \\
 \end{tabular}
@@
 
-Consider the noun \texttt{[a [b c]]}, represented as a binary tree in Figure~\ref{fig:binary-tree-example}.  Reading across in a depth-first traversal, we produce an encoding order:
+Consider the noun \texttt{[a [b c]]} (= \texttt{[a b c]}), represented as a binary tree in Figure~\ref{fig:binary-tree-example}.
+Reading across in a depth-first traversal, we produce an encoding order:
 
@@
   [(add 2 p.d) (mix 3 (lsh [0 2] q.d)) m]
+::  length encode
+++  mat
+  ~/  %mat
+  |=  a=@
+  ^-  [p=@ q=@]
+  ?:  =(0 a)
+    [1 1]
+  =+  b=(met 0 a)
+  =+  c=(met 0 b)
+  :-  (add (add c c) b)
+  (cat 0 (bex c) (mix (end [0 (dec c)] b) (lsh [0 (dec c)] a)))
 \end{lstlisting}
@@
 \begin{lstlisting}[style=listingcode]
-TODO
+++  cue
+  ~/  %cue
+  |=  a=@
+  ^-  *
+  =+  b=0
+  =+  m=`(map @ *)`~
+  =<  q
+  |-  ^-  [p=@ q=* r=(map @ *)]
+  ?:  =(0 (cut 0 [b 1] a))
+    =+  c=(rub +(b) a)
+    [+(p.c) q.c (~(put by m) b q.c)]
+  =+  c=(add 2 b)
+  ?:  =(0 (cut 0 [+(b) 1] a))
+    =+  u=$(b c)
+    =+  v=$(b (add p.u c), m r.u)
+    =+  w=[q.u q.v]
+    [(add 2 (add p.u p.v)) w (~(put by r.v) b w)]
+  =+  d=(rub c a)
+  [(add 2 p.d) (need (~(get by m) q.d)) m]
+::  length decode
+++  rub
+  ~/  %rub
+  |=  [a=@ b=@]
+  ^-  [p=@ q=@]
+  =+  ^=  c
+      =+  [c=0 m=(met 0 b)]
+      |-  ?<  (gth c m)
+      ?.  =(0 (cut 0 [(add a c) 1] b))
+        c
+      $(c +(c))
+  ?:  =(0 c)
+    [1 0]
+  =+  d=(add a +(c))
+  =+  e=(add (bex (dec c)) (cut 0 [d (dec c)] b))
+  [(add (add c c) e) (cut 0 [(add d (dec c)) e] b)]
 \end{lstlisting}
@@
 
-\section{Directed Graph Encoding}
-
-A directed graph encoding has been independently proposed twice, once by Tlon in the original Hoon codebase as a \lstinline[style=inlinecode]{+$silo} encoding and once by Sam Parker as the \lstinline[style=inlinecode]{+bulk} encoding.
-
 \section{Aligned Serialization}
@@
 
-\section{Benchmarks}
-
 \section{Conclusion}
 
+\tombstone
+
+\selectlanguage{USenglish}
+% \printbibliography
+
 \section*{Appendix A:  Python \texttt{+jam}/\texttt{+cue}}
+\subsection*{Paul Driver \patp{fodwyt-ragful}}
 
-The following is a simple Python implementation of \texttt{+jam} serialization drawn from Urbit’s auxiliary \texttt{pynoun} library.
+The following is a simple Python implementation of \texttt{+jam} serialization and \texttt{+cue} deserialization drawn from Urbit’s auxiliary \texttt{pynoun} library.\footnote{The original \texttt{pynoun} library was authored by Paul Driver \patp{fodwyt-ragful}.}
 

### 2025-11-29 mss.tex
Paginate

@@
               %% We toggle between these:
-              Manuscript submitted for review.\\
-              % {\it Urbit Systems Technical Journal} II:2 (2025):  1–46. \\
+              % Manuscript submitted for review.\\
+              {\it Urbit Systems Technical Journal} III:1 (2026):  83–101. \\
               % ~ \\
-              % Address author correspondence to \authorpatp.
+              Address author correspondence to \patp{lagrev-nocfep}.
               }}
@@
 Noun serialization is commonly used for Nock communication, both between instances like Urbit ships and with the runtime and the Unix host operating system.
-This article describes the principal convention for representing a noun in compressed form as a byte array, as well as introduces variant encodings for educational purposes.
+This article describes the principal convention for representing a noun in compressed form as a byte array and its runtime affordance.
 \end{abstract}
@@
 \footnotetext[1]{\sloppy We gratefully acknowledge the original architect of \texttt{+jam} and \texttt{+cue}, \mbox{Curtis} Yarvin \patp{sorreg-namtyv}.
-The original function, in \href{https://github.com/cgyarvin/urbit/blame/55f591594080d276fc446ceafa4afff092561613/watt/jam.watt}{\texttt{jam.watt}}, dates to \textasciitilde 2011.6.11.}
+The earliest implementation, in \href{https://github.com/cgyarvin/urbit/blame/55f591594080d276fc446ceafa4afff092561613/watt/jam.watt}{\texttt{jam.watt}}, dates to \textasciitilde 2011.6.11; Yarvin has indicated that an earlier draft no longer survives.}
 \renewcommand{\thefootnote}{\arabic{footnote}}
@@
 \pagenumbering{arabic}
-\setcounter{page}{1}
+\setcounter{page}{83}
 
@@
 \begin{tabular}{ l l r }
-  \textbf{Atom} $a$ & \textbf{Length (bits)} & \multicolumn{1}{l}{\textbf{Encoded Form} $\bar{a}$} \\
+  \textbf{Atom} $\bm{a}$ & \textbf{Length (bits)} & \multicolumn{1}{l}{\textbf{Encoded Form} $\bm{\bar{a}}$} \\
   \hline
@@
   \texttt{0xff} & 8 & \texttt{0b11111111100000000} \\
-  \texttt{0x1234} & 15 & \texttt{0b100001100100001} \\
-  & & \texttt{↳1000000000000000} \\
+  \texttt{0x1234} & 13 & \texttt{0b1001000110100} \\
+  & & \texttt{↳10000000000000} \\
 \end{tabular}
@@
   \texttt{0xff} & 8 & \texttt{0b\textcolor{lightgray}{1}1111111100000000} \\
-  \texttt{0x1234} & 15 & \texttt{0b\textcolor{lightgray}{10000}1100100001} \\
-  & & \texttt{↳1000000000000000} \\
+  \texttt{0x1234} & 13 & \texttt{0b\textcolor{lightgray}{100}1000110100} \\
+  & & \texttt{↳10000000000000} \\
 \end{tabular}
@@
 
-The primary problem with the \texttt{+fat} scheme is that the encoding of an atom is twice the bit length of the original atom.
-The next optimization is to use the bit length of the bit length, which is logarithmic in the bit length.
+The primary problem with the \texttt{+fat} scheme is that the encoding of an atom is about twice the bit length of the original atom.
+The next optimization is to use the bit length of the unary bit length, which is logarithmic in the bit length.
 
@@
 \label{tab:mat-encodings}
-\begin{tabular}{ r c c r }
-  \multicolumn{1}{l}{\textbf{Atom}} $a$ & \textbf{Length (bits)} & \textbf{Mathematics} & \multicolumn{1}{l}{\textbf{Encoded Form} $\bar{a}$} \\
+\begin{tabular}{ r c r l r }
+  \multicolumn{1}{l}{\textbf{Atom}} $\bm{a}$ & \textbf{Length (bits)} & \multicolumn{2}{c}{\hspace{2cm}\textbf{Mathematics}} & \multicolumn{1}{c}{\textbf{Encoded Form} $\bm{\bar{a}}$} \\
   \hline
-  \texttt{0b0} & $\ell = 1$ & $\underbrace{\cancel{{a_{\ell-1}}}\ldots\cancel{a_{0}}}_{\ell=0}\mathtt{1}\underbrace{\cancel{\mathtt{0}}\ldots\cancel{\mathtt{0}}}_{h=0}$ & \texttt{0b1} \\
-  \texttt{0b1} & $\ell = 1$ & $\underbrace{\mathtt{1}}_{\ell=1}\underbrace{\cancel{{\ell_{h-2}}}\ldots\cancel{\ell_{0}}}_{h-1=0}\mathtt{1}\underbrace{\mathtt{0}}_{h=1}$ & \texttt{0b110} \\
-  \texttt{0b10} & $\ell = 2$ & $\underbrace{\mathtt{10}}_{a=2}\underbrace{\mathtt{0}}_{\textrm{low bits of }\ell}\mathtt{1}\underbrace{\mathtt{00}}_{h=2}$ & \texttt{0b100100} \\
-  \texttt{0b11} & $\ell = 2$ & $\underbrace{\mathtt{11}}_{a=3}\underbrace{\mathtt{0}}_{\textrm{low bits of }\ell}\mathtt{1}\underbrace{\mathtt{00}}_{h=2}$ & \texttt{0b110100} \\
-  \texttt{0b100} & $\ell = 3$ & $\underbrace{\mathtt{100}}_{a=4}\underbrace{\mathtt{1}}_{\textrm{low bits of }\ell}\mathtt{1}\underbrace{\mathtt{00}}_{h=2}$ & \texttt{0b1001100} \\
-  \texttt{0b101} & $\ell = 3$ & $\underbrace{\mathtt{101}}_{a=5}\underbrace{\mathtt{1}}_{\textrm{low bits of }\ell}\mathtt{1}\underbrace{\mathtt{00}}_{h=2}$ & \texttt{0b1011100} \\
-  \texttt{0b1111} & $\ell = 4$ & $\underbrace{\mathtt{1111}}_{a=15}\underbrace{\mathtt{00}}_{\textrm{low bits of }\ell}\mathtt{1}\underbrace{\mathtt{000}}_{h=3}$ & \texttt{0b1111001000} \\
-  \texttt{0x70} & $\ell = 7$ & $\underbrace{\mathtt{1110000}}_{a=112}\underbrace{\mathtt{11}}_{\textrm{low bits of }\ell}\mathtt{1}\underbrace{\mathtt{000}}_{h=3}$ & \texttt{0b1110000111000} \\
+  \texttt{0b0} & $\ell = 1$ & $\underbrace{\cancel{{a_{\ell-1}}}\ldots\cancel{a_{0}}}_{\ell=0}$&$\mathtt{1}\underbrace{\cancel{\mathtt{0}}\ldots\cancel{\mathtt{0}}}_{h=0}$ & \texttt{0b1} \\
+  \texttt{0b1} & $\ell = 1$ & $\underbrace{\mathtt{1}}_{\ell=1}\underbrace{\cancel{{\ell_{h-2}}}\ldots\cancel{\ell_{0}}}_{h-1=0}$&$\mathtt{1}\underbrace{\mathtt{0}}_{h=1}$ & \texttt{0b110} \\
+  \texttt{0b10} & $\ell = 2$ & $\underbrace{\mathtt{10}}_{a=2}\underbrace{\mathtt{0}}_{\textrm{low bits of }\ell}$&$\mathtt{1}\underbrace{\mathtt{00}}_{h=2}$ & \texttt{0b100100} \\
+  \texttt{0b11} & $\ell = 2$ & $\underbrace{\mathtt{11}}_{a=3}\underbrace{\mathtt{0}}_{\textrm{low bits of }\ell}$&$\mathtt{1}\underbrace{\mathtt{00}}_{h=2}$ & \texttt{0b110100} \\
+  \texttt{0b100} & $\ell = 3$ & $\underbrace{\mathtt{100}}_{a=4}\underbrace{\mathtt{1}}_{\textrm{low bits of }\ell}$&$\mathtt{1}\underbrace{\mathtt{00}}_{h=2}$ & \texttt{0b1001100} \\
+  \texttt{0b101} & $\ell = 3$ & $\underbrace{\mathtt{101}}_{a=5}\underbrace{\mathtt{1}}_{\textrm{low bits of }\ell}$&$\mathtt{1}\underbrace{\mathtt{00}}_{h=2}$ & \texttt{0b1011100} \\
+  \texttt{0b1111} & $\ell = 4$ & $\underbrace{\mathtt{1111}}_{a=15}\underbrace{\mathtt{00}}_{\textrm{low bits of }\ell}$&$\mathtt{1}\underbrace{\mathtt{000}}_{h=3}$ & \texttt{0b1111001000} \\
+  \texttt{0x70} & $\ell = 7$ & $\underbrace{\mathtt{1110000}}_{a=112}\underbrace{\mathtt{11}}_{\textrm{low bits of }\ell}$&$\mathtt{1}\underbrace{\mathtt{000}}_{h=3}$ & \texttt{0b1110000111000} \\
+  \texttt{0x1234} & $\ell = 13$ & $\underbrace{\mathtt{1001000110100}}_{a=4660}\underbrace{\mathtt{101}}_{\textrm{low bits of }\ell}$&$\mathtt{1}\underbrace{\mathtt{0000}}_{h=4}$ & \texttt{0b100100011010010110000} \\
 \end{tabular}
@@
 
-With \texttt{+mat} encoding in hand, we can produce a proper serialization scheme for nouns by combining the type bit with the encoded atom or recursively encoding cells.
+\sloppy
+With \texttt{+mat} encoding in hand, we can produce a proper \mbox{serialization} scheme for nouns by combining the type bit with the encoded atom or recursively encoding cells.
 We call this scheme \texttt{+gel} encoding.
@@
 
-TODO opposite of gel as goo
-
 Our code implementation for \texttt{+gel} is as follows:
@@
 Whatever the pedagogical advantages of \texttt{+gel}, the algorithm is still relatively verbose.
-Nock-based operating functions have preferred in practice to use \texttt{+jam} encoding, which is a more compact run-length encoding with subtree deduplication.\footnote{Note the calculation of \patp{dozreg-toplud}, p.~TODO of this issue, that an operational Arvo instance may have up to $1.66 \times 10^{21}$ nouns, greatly reduced by structural sharing.}  \texttt{+jam} improves the basic strategy by altering the \textsc{rle} algorithm slightly and supporting internal references for noun subtrees that have already been encoded.
+Nock-based operating functions have preferred in practice to use \texttt{+jam} encoding, which is a more compact run-length encoding with subtree deduplication.\footnote{Note the calculation of \patp{dozreg-toplud}, p.~57 of this issue, that an operational Arvo instance may have up to $1.66 \times 10^{21}$ nouns, greatly reduced by structural sharing.}  \texttt{+jam} improves the basic strategy by altering the \textsc{rle} algorithm slightly and supporting internal references for noun subtrees that have already been encoded.
 
@@
 
-A Python implementation of \texttt{+jam} is included in Appendix~A.
+\sloppy
+A Python implementation of \texttt{+jam} is included in the \mbox{appendix}.
 
@@
 
-Likewise, a Python implementation of \texttt{+cue} is included in Appendix~A.
+Likewise, a Python implementation of \texttt{+cue} is included in the \mbox{appendix}.
 
@@
 
-\section{Aligned Serialization}
+% \section{Aligned Serialization}
 
-One of the advantages of \texttt{+jam} is its compactness.
-However, this comes at the cost of speed, since bit-level operations are required to \texttt{+cue} the noun back from its serialized form.
-If a slightly larger size is acceptable, a byte-aligned serialization could facilitate certain kinds of external inspection without requiring deserialization.
-(For instance, a byte-aligned head tag could be read for a rapid decision without needing to \texttt{+cue} the entire noun.)
+% One of the advantages of \texttt{+jam} is its compactness.
+However, this comes at the cost of decoding speed, since bit-level operations are required to \texttt{+cue} the noun back from its serialized form.
+If a larger size is acceptable, a byte-aligned serialization could facilitate certain kinds of external inspection without requiring any deserialization at all.
+For instance, a byte-aligned head tag could be read for a rapid decision without needing to \texttt{+cue} the entire noun.
+Byte-aligned values are straightforward for systems languages like C (Vere) and Rust (NockVM) to examine and manipulate efficiently, so there is some theoretical attraction to preferring them.
 
-We propose a strategy to modify \texttt{+jam} to align to bytes by padding the length of entries to the nearest byte boundary and marking the distance with a clever binary scheme rather than simply unary.
+% We propose a strategy to modify \texttt{+jam} to align to bytes by padding the length of entries to the nearest byte boundary and marking the distance with a clever binary scheme rather than simply unary.
 This approach, called \texttt{+honey}, aims to balance compactness and speed for certain use cases while retaining a large degree of conceptual backwards compatibility.
@@
 
-There are two fundamental issues for byte alignment:  atoms and lengths.
+% As outlined above, there are two fundamental issues for serialization:  types and lengths.
 Atoms can be padded with leading zeros to the nearest byte boundary without changing their value.
-Lengths, however, require a new encoding scheme to compensate for the adjustment in expected bit widths.
+Lengths, however, require a tweak to their encoding scheme to compensate for the adjustment in expected bit widths.
 %We propose a variable-length encoding for lengths that uses the high bit of each byte to indicate whether more bytes follow.
@@
 
+% \paragraph*{\texttt{+honey} Serialization Algorithm}
+% \begin{itemize}
+% \item  \textbf{Encode}.
+%   \begin{itemize}
+%     \item  If the noun is an atom, write \texttt{0} followed by the \texttt{+mat} encoding of the atom, padded to the nearest byte boundary.
+%     \item  If the noun is a cell, check if it has been previously encoded.
+%       \begin{itemize}
+%         \item  If it has, write \texttt{11} followed by the byte-aligned encoding of the index where it was first encoded.
+%         \item  If it has not, write \texttt{01} followed by the \texttt{+honey} encoding of the head and then the \texttt{+honey} encoding of the tail.  Store the current index for future reference.
+%       \end{itemize}
+%   \end{itemize}
+% \item  \textbf{Decode}.  Read the first byte.
+%   \begin{itemize}
+%     \item  If it is \texttt{0}, read the next bytes as a \texttt{+mat} encoded atom, removing any padding.
+%     \item  If it is \texttt{1}, read the next byte.
+%       \begin{itemize}
+%         \item  If it is \texttt{0}, recursively read the next bytes as the head and then the tail.
+%         \item  If it is \texttt{1}, read the next bytes as a byte-aligned encoded index and retrieve the cached subtree.
+%       \end{itemize}
+%   \end{itemize}
+% \end{itemize}
+
+% A \texttt{+honey}-encoded noun has the following structure:
+
+% \begin{lstlisting}[style=listingcode]
+%   T.LLLL.HHHH.HHHH.HHHH.HHHH.HHHH.HHHH...
+% \end{lstlisting}
+
+% \noindent
+% where blah blah blah.  Examples of \texttt{+honey} encodings are given in Table~\ref{tab:honey-encodings}.
+
+% \begin{table}[ht]
+% \centering
+% \caption{Example noun encodings (\texttt{+honey} encoding).
+\texttt{.} dot delimits byte fields for readability.
+The atom \texttt{\%head} is a text tag encoded as \texttt{0x68.0x65.0x61.0x64} in \textsc{ascii}.}
+% \label{tab:honey-encodings}
+% \begin{tabular}{ r r }
+%   \multicolumn{1}{l}{\textbf{Noun}} & \multicolumn{1}{l}{\textbf{Encoded Form}} \\
+%   \hline
+%   \texttt{[0 0]} & \texttt{0x00.0x01.0x00.0x01} \\
+%   \texttt{[1 0]} & \texttt{0x00.0x06.0x00.0x01} \\
+%   \texttt{[2 1 0]} & \texttt{0x00.0x06.0x00.0x01.0x04.0x01} \\
+%   \texttt{[[2 3] 1 0]} & \texttt{0x00.0x0C.0x00.0x01.0x06.0x04.0x01.0x01} \\
+%   \texttt{[\%head 42]} & \texttt{...} \\
+% \end{tabular}
+% \end{table}
+
+% This scheme allows for efficient byte-aligned serialization and deserialization of nouns while retaining the benefits of subtree deduplication from \texttt{+jam}.
+Whether \texttt{+honey} or a similar byte-aligned scheme is practical for runtime use remains to be seen.
+
 \section{Conclusion}
 
-\tombstone
+Noun serialization is a fundamental aspect of Urbit's architecture, enabling efficient communication of untyped data structures.
+Starting from basic run-length encoding, we have explored progressively more sophisticated schemes culminating in \texttt{+jam}, which balances compactness and efficiency through subtree deduplication.
+Further extensions like \texttt{newt}
+%and \texttt{+honey}
+illustrate the adaptability of serialization strategies to meet evolving runtime requirements.
+As Nock-based solid-state computing continues to evolve, noun serialization will remain an important foundation for data interchange and persistence.\tombstone
 
@@
 
-\section*{Appendix A:  Python \texttt{+jam}/\texttt{+cue}}
+\section*{Appendix:  Python \texttt{+jam}/\texttt{+cue}}
 \subsection*{Paul Driver \patp{fodwyt-ragful}}
@@
 # The Cell class represents an ordered pair of two nouns.
-
 def jam_to_stream(n: noun, out: BitArray):
     """jam but put the bits into a stream
-
     >>> s = BitArray()
@@
         refs[a] = cur
-
     def mat(i: int):
@@
     """cue but read the bits from a stream
-
     >>> s = BitArray('0b01')

### 2025-12-01 mss.tex
Fix +gel.

@@
 
+The reference implementation of \texttt{+mat} encoding in Hoon is as follows:
+
+\begin{lstlisting}[style=listingcode]
+++  mat
+  ~/  %mat
+  |=  a=@
+  ^-  [p=@ q=@]
+  ?:  =(0 a)
+    [1 1]
+  =+  b=(met 0 a)
+  =+  c=(met 0 b)
+  :-  (add (add c c) b)
+  (cat 0 (bex c) (mix (end [0 (dec c)] b)
+   ↳                  (lsh [0 (dec c)] a)))
+\end{lstlisting}
+
 \begin{landscape}
@@
   \hline
+  \texttt{0} & \texttt{0b0.1.0} \\
   \texttt{[0 0]} & \texttt{0b1.0.1.0.1} \\
-  \texttt{[1 0]} & \texttt{0b110.0.1.0.1} \\
-  \texttt{[2 1 0]} & \texttt{0b110.0.1.0.1.100100.0.1} \\
-  \texttt{[[2 3] 1 0]} & \texttt{0b110.0.1.0.1.110100.0.100100.0.1.1} \\
+  \texttt{[1 0]} & \texttt{0b1.0.110.0.1} \\
+  \texttt{[2 1 0]} & \texttt{0b1.0.110.0.1.100100.0.1} \\
+  \texttt{[[2 3] 1 0]} & \texttt{0b1.0.110.0.1.110100.0.100100.0.1.1} \\
 \end{tabular}
@@
 
-Our code implementation for \texttt{+gel} is as follows:
+\newpage
+Our code implementation for \texttt{+gel} encoding is as follows:
 
 \begin{lstlisting}[style=listingcode]
-!:  |%
 ++  gel
-  =gel !:  |=  a=*
+  |=  a=*
   ^-  @
-  =+  l=0
   =+  b=0
-  =<  -
-  |-
-  ?^  a
-    =+  lv=$(a -.a)
-    =+  rv=$(a +.a)
-    =+  [c l]=(mash rv lv)
-    [(con (lsh [0 1] c) 0b1) +(l)]
-  ?:  =(0 a)  [0b10 4]
-  :: need another mash in here for unary length
-  =+  [c l]=(mash [a l] [b +((met 0 b))])
-  [(con (lsh [0 1] c) 0b0) +(l)]
-:: length of atom in unary
-++  len
-  |=  a=@
-  ^-  @
-  (fil 0 (met 0 a) 0b1)
-:: mash two atoms together
-++  mash
-  |=  [a=[p=@ l=@] b=[p=@ l=@]]
-  ^-  [c=@ l=@]
-  :-  (con (lsh [0 l.b] p.a) p.b)
-  (add l.a l.b)
---
+  =<  q
+  |-  ^-  [p=@ q=@]
+  ?:  ?=(@ a)
+    =+  d=(mat a)
+    [(add 1 p.d) (lsh 0 q.d)]
+  =>  .(b (add 1 b))
+  =+  d=$(a -.a)
+  =+  e=$(a +.a, b (add b p.d))
+  :-  (add 1 (add p.d p.e))
+  (mix 1 (lsh [0 1] (cat 0 q.d q.e)))
 \end{lstlisting}
+
+\noindent
+Decoding is left as an exercise to the interested reader.
 
@@
   [(add 2 p.d) (mix 3 (lsh [0 2] q.d)) m]
-::  length encode
-++  mat
-  ~/  %mat
-  |=  a=@
-  ^-  [p=@ q=@]
-  ?:  =(0 a)
-    [1 1]
-  =+  b=(met 0 a)
-  =+  c=(met 0 b)
-  :-  (add (add c c) b)
-  (cat 0 (bex c) (mix (end [0 (dec c)] b) (lsh [0 (dec c)] a)))
 \end{lstlisting}

### 2025-12-27 mss.tex
Post
Fix
Revise

@@
 \usepackage{pdflscape}
+\usepackage{array}  % for table column formatting
 
@@
 \centering
-\caption{Example atom encodings with length information appended as trailing zeros (unnamed encoding).}
+\caption{Example atom encodings with length information appended as trailing zeros (unnamed encoding).  Elided digits are shown in light gray for clarity.}
 \label{tab:atom-encodings}
@@
   \hline
-  \texttt{0b0} & 1 & \texttt{0b1} \\
+  \texttt{0b0} & 1 & \texttt{0b\textcolor{lightgray}{0}1} \\
   \texttt{0b1} & 1 & \texttt{0b110} \\
@@
   \texttt{0xff} & 8 & \texttt{0b11111111100000000} \\
-  \texttt{0x1234} & 13 & \texttt{0b1001000110100} \\
-  & & \texttt{↳10000000000000} \\
+  \texttt{0x1234} & 13 & \llap{\texttt{0b100100011010010000000000000}} \\
 \end{tabular}
@@
   \texttt{0xff} & 8 & \texttt{0b\textcolor{lightgray}{1}1111111100000000} \\
-  \texttt{0x1234} & 13 & \texttt{0b\textcolor{lightgray}{100}1000110100} \\
-  & & \texttt{↳10000000000000} \\
+  \texttt{0x1234} & 13 & \llap{\texttt{0b\textcolor{lightgray}{100}100011010010000000000000}} \\
+  % \texttt{0x1234} & 13 & \texttt{0b\textcolor{lightgray}{100}100011010010000000000000} \\
+  % \texttt{0x1234} & 13 & \texttt{0b\textcolor{lightgray}{100}1000110100} \\
+  % & & \texttt{↳10000000000000} \\
 \end{tabular}
@@
 
-Consider the noun \texttt{[a [b c]]} (= \texttt{[a b c]}), represented as a binary tree in Figure~\ref{fig:binary-tree-example}.
+Consider the noun \texttt{[4 [0 1]]} (= \texttt{[4 0 1]}), represented as a binary tree in Figure~\ref{fig:binary-tree-example}.
 Reading across in a depth-first traversal, we produce an encoding order:
@@
   \hline
-  \texttt{0} & \texttt{0b0.1.0} \\
+  \texttt{0} & \texttt{0b\textcolor{lightgray}{0}.1.0} \\
   \texttt{[0 0]} & \texttt{0b1.0.1.0.1} \\
@@
 \noindent
-Decoding is left as an exercise to the interested reader.
+Decoding is left as an exercise to the motivated reader.
 
@@
 % \printbibliography
+\newpage
 
@@
 \subsection*{Paul Driver \patp{fodwyt-ragful}}
+% add to table of contents
+\addcontentsline{toc}{section}{Appendix:  Python \texttt{+jam}/\texttt{+cue}}
 
-The following is a simple Python implementation of \texttt{+jam} serialization and \texttt{+cue} deserialization drawn from Urbit’s auxiliary \texttt{pynoun} library.\footnote{The original \texttt{pynoun} library was authored by Paul Driver \patp{fodwyt-ragful}.}
+The following is a simple Python implementation of \texttt{+jam} serialization and \texttt{+cue} deserialization drawn from Urbit’s auxiliary \texttt{pynoun} library.%\footnote{The original \texttt{pynoun} library was authored by Paul Driver \patp{fodwyt-ragful}.}
 
@@
         refs[a] = cur
+
     def mat(i: int):
